---
title: "Basic SQL Tutorial"
output: 
  html_document:
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This tutorial provides a rundown on SQL commands to query relational databases. The content of this tutorial is inspired by IBM's [Databases and SQL for Data Science with Python](https://www.coursera.org/learn/sql-data-science) on Coursera. The tutorial assumes prior knowledge on relational database concepts

## Dataset

The data for this tutorial are from Chicago's Data Portal, consisting of the following:

1. [Socioeconomic Indicators in Chicago](https://data.cityofchicago.org/Health-Human-Services/Census-Data-Selected-socioeconomic-indicators-in-C/kn9c-c2s2)
    * This is a cross-sectional (single-year) data containing socioeconomic indicators (**Crowding, Poverty, Unemployment, Age distribution, Income**) for each community area in Chicago
2. [Chicago Public Schools](https://data.cityofchicago.org/Education/Chicago-Public-Schools-Progress-Report-Cards-2011-/9xs2-f89t)
    * This is a cross-section (single-year) data containing school performance indicators (**Health Certification, Safety Score, College Enrollment**) for each school in each comunity area in Chicago
3. [Chicago Crime Data](https://data.cityofchicago.org/Public-Safety/Crimes-2001-to-present/ijzp-q8t2)
    * This is a panel (multi-year) data containing crime information (**Crime type, Arrest indicator**) for each crime that occured in Chicago

### Download (csv data)

As the data from the three sources are very large, a smaller subset is used for this tutorial instead, with more database-friendly column names

1. [Census Dataset](https://cf-courses-data.s3.us.cloud-object-storage.appdomain.cloud/IBMDeveloperSkillsNetwork-DB0201EN-SkillsNetwork/labs/FinalModule_Coursera/data/Census_Data_-_Selected_socioeconomic_indicators_in_Chicago__2008___2012-v2.csv) (read as **CENSUS_DATA**)
2. [Chicago Public School](https://cf-courses-data.s3.us.cloud-object-storage.appdomain.cloud/IBMDeveloperSkillsNetwork-DB0201EN-SkillsNetwork/labs/FinalModule_Coursera/data/Chicago_Public_Schools_-_Progress_Report_Cards__2011-2012-v3.csv) (read as **CHICAGO_PUBLIC_SCHOOLS**)
3. [Chicago Crime Data](https://cf-courses-data.s3.us.cloud-object-storage.appdomain.cloud/IBMDeveloperSkillsNetwork-DB0201EN-SkillsNetwork/labs/FinalModule_Coursera/data/Chicago_Crime_Data-v2.csv) (**CHICAGO_CRIME_DATA**)

## Reading data

As there is no live database for this tutorial, the datasets will have to be read in as local dataframes. We will also use the **sqldf** library to practice SQL queries
```{r read, warning = FALSE,message = FALSE}
CENSUS_DATA = read.csv('Census_Data_-_Selected_socioeconomic_indicators_in_Chicago__2008___2012-v2.csv')
CHICAGO_PUBLIC_SCHOOLS = read.csv('Chicago_Public_Schools_-_Progress_Report_Cards__2011-2012-v3.csv')
CHICAGO_CRIME_DATA = read.csv('Chicago_Crime_Data-v2.csv')
library(sqldf)
library(DT)
```

Let us take a look at some of the variables in the data that we have

### Census preview
```{r preview,echo=FALSE}
datatable(head(CENSUS_DATA),options=list(scrollX=T))
```

### School preview
```{r preview2,echo=FALSE}
datatable(head(CHICAGO_PUBLIC_SCHOOLS),options=list(scrollX=T))
```

### Crime preview
```{r preview3,echo=FALSE}
datatable(head(CHICAGO_CRIME_DATA),options=list(scrollX=T))
```

## Basic SQL Statements

### SELECT
SELECT is used to retrieve data from a table. The syntax is __SELECT * FROM TABLE__. This will query the entire table
```{r SELECT}
syntax = 'SELECT * FROM CENSUS_DATA'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

#### SELECT specific columns
We can also use SELECT on specific columns for a table. The syntax is **SELECT COL_A, COL_B FROM TABLE**. This will query only the specified columns for the entire table
```{r SELECTspecific}
syntax = 'SELECT COMMUNITY_AREA_NUMBER, COMMUNITY_AREA_NAME FROM CENSUS_DATA'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

#### LIMIT
LIMIT is used to restrict the number of rows from a data base. The syntax is __SELECT * FROM TABLE LIMIT NUMBER__, where NUMBER is the number of rows required
```{r SELECTlimit}
syntax = 'SELECT COMMUNITY_AREA_NUMBER, COMMUNITY_AREA_NAME FROM CENSUS_DATA LIMIT 3'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### WHERE
WHERE clauses are used to restrict the selection to subsets only. The syntax is _SELECT * FROM TABLE_ __WHERE PREDICATE__. For instance, we want communities with *PER_CAPITA_INCOME >= 50000*
```{r WHERE}
syntax = 'SELECT COMMUNITY_AREA_NUMBER, COMMUNITY_AREA_NAME, PER_CAPITA_INCOME FROM CENSUS_DATA WHERE PER_CAPITA_INCOME>=50000'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```
More conditions are available here

|         Condition|  Syntax|
|:-----------------|----:|
|Equal to          | =   |
|Greater than      | >   |
|Less than         | <   |
|Greater than or equal to    | >=|
|Less than or equal to | <=|
|Not equal to      | <>|

### Other useful expressions

#### COUNT
COUNT gives the number of rows that match a query. For instance, we want to know how many communities have *PER_CAPITA_INCOME >= 50000*. We can also rename the column name to something we want using **AS**. This will give 6, which is the same number of rows as the previous example
```{r COUNT}
syntax = 'SELECT COUNT(*) AS NUMBER FROM CENSUS_DATA WHERE PER_CAPITA_INCOME>=50000'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

#### DISTINCT
DISTINCT removes duplicates from the query. For instance, we want to find the unique types of crime from CHICAGO_CRIME_DATA that occur at GAS STATION locations
```{r DISTINCT}
syntax = 'SELECT DISTINCT(PRIMARY_TYPE) AS CRIME_TYPE, LOCATION_DESCRIPTION FROM CHICAGO_CRIME_DATA WHERE LOCATION_DESCRIPTION = "GAS STATION"'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```


## Advanced SQL Statements

### LIKE string pattern

String patterns are useful when we do not know the exact predicate we need. The syntax is _SELECT * FROM TABLE WHERE_ __PREDICATE LIKE '%B'__  For instance, we want all Community Areas that start with B
```{r LIKE}
syntax = 'SELECT COMMUNITY_AREA_NAME FROM CENSUS_DATA WHERE COMMUNITY_AREA_NAME LIKE "B%"'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### AND ranges

We can also select results where the predicate is between a range. Suppose we want Community Areas where HARDSHIP_INDEX is between 1 to 10
```{r AND}
syntax = 'SELECT COMMUNITY_AREA_NAME, HARDSHIP_INDEX FROM CENSUS_DATA WHERE HARDSHIP_INDEX>=1 AND HARDSHIP_INDEX<=10'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

#### BETWEEN AND
Alternatively, we can use a __BETWEEN...AND__ statement for the same query
```{r BETWEEN}
syntax = 'SELECT COMMUNITY_AREA_NAME, HARDSHIP_INDEX FROM CENSUS_DATA WHERE HARDSHIP_INDEX BETWEEN 1 AND 10'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### OR sets
If we want to retrieve data using a set of values, the OR operator comes in handy. It follows the same syntax as AND. If we want a community with 6 or 10 for HARDSHIP_INDEX, we can do the following:
```{r OR}
syntax = 'SELECT COMMUNITY_AREA_NAME, HARDSHIP_INDEX FROM CENSUS_DATA WHERE HARDSHIP_INDEX = 6 OR HARDSHIP_INDEX = 10'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

#### IN
Similarly, we can also use the IN operator, which is more efficient to type if we have many values to choose from
```{r IN}
syntax = 'SELECT COMMUNITY_AREA_NAME, HARDSHIP_INDEX FROM CENSUS_DATA WHERE HARDSHIP_INDEX IN (6,7,10)'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### ORDER BY
This clause is used for sorting results. The syntax is _SELECT * FROM TABLE_ **ORDER BY COL** where COL is the column name, or column number. To order in descending order, add **DESC** after specifying the column. For instance, if we want the top 5 communities by HARDSHIP_INDEX, we can do the following
```{r ORDER}
syntax = 'SELECT COMMUNITY_AREA_NAME, HARDSHIP_INDEX FROM CENSUS_DATA ORDER BY 2 DESC LIMIT 5'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### GROUP BY
The GROUP BY clause groups results into subsets that have matching values for one (or more) columns. The syntax is _SELECT * FROM TABLE_ __GROUP BY COL__, where COL is the column to be grouped For instance, we want to count the number of crimes that occured at each different location
```{r GROUP}
syntax = 'SELECT LOCATION_DESCRIPTION, COUNT(LOCATION_DESCRIPTION) FROM CHICAGO_CRIME_DATA GROUP BY LOCATION_DESCRIPTION'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

#### HAVING restriction
A restriction can be applied immediately after a GROUP BY clause using HAVING. The syntax is _SELECT * FROM TABLE GROUP BY COL_ __HAVING PREDICATE__. For instance, we only want crime locations where there are at least 20 crimes
```{r HAVING}
syntax = 'SELECT LOCATION_DESCRIPTION, COUNT(LOCATION_DESCRIPTION) FROM CHICAGO_CRIME_DATA GROUP BY LOCATION_DESCRIPTION HAVING COUNT(LOCATION_DESCRIPTION) >= 20'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```


### Built-in functions
#### Aggregate functions

|         Function|  Result|
|:-----------------|----:|
|SUM         | Add all values of column   |
|MIN         | Return the minimum of column   |
|MAX         | Return maximum of column   |
|AVERAGE     | Return average of column|

Example: We want to find the average safety score of schools by their community area
```{r FUNCTIONS}
syntax = 'SELECT AVG(SAFETY_SCORE), COMMUNITY_AREA_NAME FROM CHICAGO_PUBLIC_SCHOOLS GROUP BY COMMUNITY_AREA_NAME'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

#### Scaler and String functions
|         Function|  Result|
|:-----------------|----:|
|ROUND         | Round all values in column   |
|LENGTH         | Return length of all values in column   |
|UCASE         | Return uppercase of all values in column   |
|LCASE     | Return lowercase of all values in column|

Apart from being applies in the SELECT clause, functions can also be applied in the left side of WHERE/HAVING clause. For instance, we want to filter the community areas from above such that they have an average safety score of at least 80

```{r SCALER}
syntax = 'SELECT AVG(SAFETY_SCORE), COMMUNITY_AREA_NAME FROM CHICAGO_PUBLIC_SCHOOLS GROUP BY COMMUNITY_AREA_NAME HAVING AVG(SAFETY_SCORE) >= 80'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

## Sub-Queries and Nested selects

### Using Sub-Queries as a predicate
Aggregate functions such as AVG() cannot be applied on the right side of a WHERE clause. If we want to query for schools with an above average safety score, the following query will fail
```{r fail}
syntax = 'SELECT SAFETY_SCORE, NAME_OF_SCHOOL FROM CHICAGO_PUBLIC_SCHOOLS WHERE SAFETY_SCORE > AVG(SAFETY_SCORE)'
```
A sub-query is needed to solve this. This means adding a query as the right side of a WHERE clause. THe following query allows us to find schools with an above average safety score
```{r sq predicate}
syntax = 'SELECT SAFETY_SCORE, NAME_OF_SCHOOL FROM CHICAGO_PUBLIC_SCHOOLS WHERE SAFETY_SCORE > (SELECT AVG(SAFETY_SCORE) FROM CHICAGO_PUBLIC_SCHOOLS)'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### Using Sub-Queries in a new column
Similar to the previous example, aggregate functions also cannot be applied as a new column if there is no aggregation occurring. For example, if we want to compare the safety score of each school to the average safety score, the following query will fail
```{r fail2}
syntax = 'SELECT SAFETY_SCORE, AVG(SAFETY_SCORE), NAME_OF_SCHOOL FROM CHICAGO_PUBLIC_SCHOOLS WHERE SAFETY_SCORE > AVG(SAFETY_SCORE)'
```
Instead, a sub-query like the following will have to be used
```{r sq new col}
syntax = 'SELECT SAFETY_SCORE, (SELECT AVG(SAFETY_SCORE) FROM CHICAGO_PUBLIC_SCHOOLS) AS AVG_SAFETY_SCORE, NAME_OF_SCHOOL FROM CHICAGO_PUBLIC_SCHOOLS'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### Using Sub-Queries as a Derived Table
As derived tables, sub-queries can be useful when it comes to dealing with multiple tables. Briefly, it involves turning a sub-query into an entirely new table to be queried. The syntax is _SELECT * FROM_ __(SELECT COLA, COLB FROM TABLE)__ _AS NEWTABLE_. The usefulness is not visible whend dealing with single tables
```{r sq derived}
syntax = 'SELECT * FROM (SELECT NAME_OF_SCHOOL, SAFETY_SCORE FROM CHICAGO_PUBLIC_SCHOOLS) AS SCHOOL'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```


## Working with multiple tables
### Via subqueries
In CHICAGO_CRIME, there is a field called COMMUNITY_AREA_NUMBER, but there is no COMMUNITY_AREA_NAME. In order to filter for names, we can utilise the CENSUS_DATA table, which contains both COMMUNITY_AREA_NUMBER and COMMUNITY_AREA_NAME. Suppose we are interested in finding crimes that happen only in Lincoln Park. We can then use the following sub-query
```{r multi sub}
syntax = 'SELECT CASE_NUMBER, COMMUNITY_AREA_NUMBER FROM CHICAGO_CRIME_DATA WHERE COMMUNITY_AREA_NUMBER IN (SELECT COMMUNITY_AREA_NUMBER FROM CENSUS_DATA WHERE COMMUNITY_AREA_NAME = "Lincoln Park")'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```

### Via implicit joins
We will not be exploring full joins as they tend to give a result set containing more rows than in both tables individually. To limit a join result, we can specify a clause to matching the 2 joining columns. Note that we can use __aliases__ for table names

We can also add columns from the second table (if we are interested to see the community area where the crime occured), by specifing the __aliases and columns__ in select
```{r implicit}
syntax = 'SELECT  A.CASE_NUMBER, A.COMMUNITY_AREA_NUMBER, B.COMMUNITY_AREA_NAME FROM CHICAGO_CRIME_DATA AS A, CENSUS_DATA AS B WHERE A.COMMUNITY_AREA_NUMBER = B.COMMUNITY_AREA_NUMBER'
df = sqldf(syntax)
datatable(df,options=list(scrollX=T))
```
There are no more duplicates, and the crimes that do not have a community area number assigned are also not in the new implicit join

## Conclusion
I hope this tutorial has been useful! While SQL is a querying language and is not used for modeling, it is an important skill to learn for data scientists as obtaining data from databases is always the first step for any data science project. SQL is needed to communicate with databases, and is becoming standard use in data science




